### 2018-06-04 express2018/part2.tex
Updates following Sangiorgi's suggestions

We will not repeat all definitions and theorems of CCS again by their
formalized versions, nor we talk about the changes and improvements
comparing with previous CCS work by Nesi. Instead, we just focus on our
targeting theorems. However, the unique solution of (rooted)
contraction theorem stood at the end of a chain of and
concepts and results, it's
impossible to describe only the final statements and it formal proofs without
describing all dependent concepts first.  On the other side, it's very
hard to convince audience the novelty in a proof
engineering work, especially when there exists multiple formalizations
 in the same area. After all, the whole interactive theorem proving (ITP)
technique is just a translation of definitions and hand-made
proofs from their informal representations to their formal versions.

Unlike complicated large programs (e.g. operating systems) which
usually contain countless of bugs even after years of updates, a huge
proof engineering work can easily claim its
\emph{correctness} without concerning other aspects (performance,
quality of definitions, easiness of reading by human, etc.)
Yet still different formalizations (of the same theory) done by
different persons could be very different, due to the great
flexibilities on choices made during the design of types, terms and
definitions. It's also quite common to break the original long proofs to
small pieces. For the same theorem, good and bad-killed proof engineers
may result into very different proof steps with quite different
lenths.\footnote{But it's not true that a shorter proof is always
better, and a shorted proof may not be executed faster if it heavily
depends on automatic proof searching tools such as Meson and Metis. On
the other side, longer proof without automatic proof searching can be
easier understood by students who are learning the theory being
formalized. Indeed, we have tried to follow the informal proof steps literally, and only used
automatic proof searching techniques on obvious small proof goals.}
Roughly speaking, there're relative ``good'' and ``bad'' designs and
considerations, but if they all \emph{correctly} formalized the same theory, few people
care about their differences, unless they want to extend existing work
to verify their new developments.

In the rest of this section, we're not going to repeat any proof steps
as they're essentially the same with the informal proofs. Instead, we
mostly focus on the key definitions, i.e. the formal
representation of various concepts, plus brief descriptions on
important stage theorem expressed in these definitions.

@@

If we drop this classical assumption, then the proof becomes much
harder, as given a van Glabbek's paper. Beside, it requires that CCS
syntax must support infinite sums of processes, and this ``infinity''
will go to arbitrary ordinals. Unfortunately, this is beyond the
ability of HOL (or simple type theory). What we can prove by following
the same idea is the following version for finite state CCS, i.e. the
set of all reachable processes is finite:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{finite_state} \HOLFreeVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{finite_state} \HOLFreeVar{q} \HOLSymConst{\HOLTokenImp{}}
   (\HOLFreeVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLFreeVar{q} \HOLSymConst{\HOLTokenEquiv{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLFreeVar{p} \HOLSymConst{+} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLFreeVar{q} \HOLSymConst{+} \HOLBoundVar{r})
\end{alltt}

>>>>

We will not repeat all definitions and theorems of CCS again by their
formalized versions. Instead, we just focus on several highlights in
this work, i.e.
\begin{enumerate}
\item The use of recursion operators in formalizing process constants
  and variables;
\item The formalization of semantics context in $\lambda$-expressions and the theory of
  (pre)congruence for CCS;
\item The definition and uses of trace in the proof of unique solution of
  contractions theorem;
\item The proof of hard version of ``coarsest congruence
  contained in $\approx$'' theorem.
\end{enumerate}

@@

If we drop this classical assumption, then the proof becomes much
harder, as given a van Glabbek's paper. The most important
intermediate result we have proved here, is the following lemma:
\begin{alltt}
\HOLTokenTurnstile{} (\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}.
        \HOLConst{STABLE} \HOLBoundVar{k} \HOLSymConst{\HOLTokenConj{}} (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p\sp{\prime}} \HOLBoundVar{u}. \HOLFreeVar{p} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}(\HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k})) \HOLSymConst{\HOLTokenConj{}}
        \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{q\sp{\prime}} \HOLBoundVar{u}. \HOLFreeVar{q} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \HOLBoundVar{q\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}(\HOLBoundVar{q\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k})) \HOLSymConst{\HOLTokenImp{}}
   (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLFreeVar{p} \HOLSymConst{+} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLFreeVar{q} \HOLSymConst{+} \HOLBoundVar{r}) \HOLSymConst{\HOLTokenImp{}}
   \HOLFreeVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLFreeVar{q}
\end{alltt}
It roughly says, for any two processes $p$ and $q$, if we can find
another stable (no $\tau$-transitions) process $k$ which is not weak bisimilar to any transition of
$p$ and $q$, then the hard part of our main theorem is proved. In
practice, once two processes were given, it's not hard to find
such a process, but the arbitrariness of $p$ and $q$ made this result
extremely hard to prove. The method given by van Glabbek requires a
construction of arbitrarily non-bisimilar processes called ``Klop
processes'':
\begin{definition}{(Klop processes)}
For each ordinal $\lambda$, and an arbitrary chosen non-$\tau$ action $a$,
define a CCS process $k_\lambda$ as follows:
\begin{enumerate}
\item $k_0 = 0$,
\item $k_{\lambda+1} = k_\lambda + a.k_\lambda$ and
\item for $\lambda$ a limit ordinal, $k_\lambda = \sum_{\mu < \lambda}
  k_\mu$, meaning that $k_\lambda$ is constructed from all graphs
  $k_\mu$ for $\mu < \lambda$ by identifying their root.
\end{enumerate}
\end{definition}
It's not hard to prove that, all $k_i$ are non-bisimilar (not only
weakly but also strongly). The idea is, for any two processes, the union of sets of their all
transitions cannot be arbitrarily large: it has to be limited by an
ordinal number. But above contruction can be arbitrarily large, so
there always exists a Klop process which can be used to satisfy above
lemma (and finish the proof).

However, this goes beyond HOL's expressivity to define above process, mostly
because there's no way to express infinite ``sums'' in CCS
datatype.\footnote{There're actually several ways to modify the
  datatype to support infinite sums, but none of these infinity can be
arbitrary. The full version of ordinal theory cannot be expressed in HOL.}
What we can do is to eliminate the last branch with limiting
ordinals, and define Klop processes only on finite cases:
\begin{alltt}
\HOLConst{KLOP} \HOLFreeVar{a} \HOLNumLit{0} \HOLSymConst{=} \HOLConst{nil}
\HOLConst{KLOP} \HOLFreeVar{a} (\HOLConst{SUC} \HOLFreeVar{n}) \HOLSymConst{=} \HOLConst{KLOP} \HOLFreeVar{a} \HOLFreeVar{n} \HOLSymConst{+} \HOLConst{label} \HOLFreeVar{a}\HOLSymConst{..}\HOLConst{KLOP} \HOLFreeVar{a} \HOLFreeVar{n}\hfill[KLOP_def]
\end{alltt}
But this means we much assume the two processes $p$ and $q$ are
finite-state, i.e. their corresponding LTS graphs have only finite
states. Choosing an element from a countable infinite set of processes
to make sure it's not bisimiar with a given set of processes, this is
essentially a pure set-theoretic problem, as formalized and proved in
the following lemma:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{equivalence} \HOLFreeVar{R} \HOLSymConst{\HOLTokenImp{}}
   \HOLConst{FINITE} \HOLFreeVar{A} \HOLSymConst{\HOLTokenConj{}} \HOLConst{INFINITE} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}}
   (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{x} \HOLBoundVar{y}. \HOLBoundVar{x} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{y} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{x} \HOLSymConst{\HOLTokenNotEqual{}} \HOLBoundVar{y} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{R} \HOLBoundVar{x} \HOLBoundVar{y}) \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}. \HOLBoundVar{k} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{n}. \HOLBoundVar{n} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{A} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{R} \HOLBoundVar{n} \HOLBoundVar{k}
\end{alltt}

With all these results, finally we can prove the ``coarsest congruence
contained in $\approx$'' theorem for finite-state CCS:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{finite_state} \HOLFreeVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{finite_state} \HOLFreeVar{q} \HOLSymConst{\HOLTokenImp{}}
   (\HOLFreeVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLFreeVar{q} \HOLSymConst{\HOLTokenEquiv{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLFreeVar{p} \HOLSymConst{+} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLFreeVar{q} \HOLSymConst{+} \HOLBoundVar{r})
\end{alltt}
